Blocks of extended ancestry around selected loci in hybrid zones

### 2016-03-01 drafts/hybrid_zone.tex
typo hunt

While for much of the genome gene flow will eventually erase 
both geographic clines in ancestry, and strong correlations in ancestry,
subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. 
% Variation in \yb{ the action of } selection against migrant alleles within the genome results in genome-wide heterogeneity in cline width, as the frequency of foreign ancestry surrounding targets of selection is reduced \citep{Barton1979}. 
% While these patterns of co-ancestry have been described in multi-locus systems, the \yb{expected length of ancestral tracts around selected loci in admixing populations remains poorly understood}\alisa{, especially in the geographic context of hybrid zones.}
\yb{While empirical patterns of the block length distribution in hybrid zones have rarely been used to inform the history of selection,}  \alisa{the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection.

@@

We consider two isolated populations
--- labeled \emph{Species $A$} and \emph{Species $B$} ---  
that came into contact $T$ generations in the past.  
We will say that species $A$ was initially on the ``left'' of the zone of contact, which corresponds to spatial positions $x<0$.  
After contact, the populations live across continuous geographical space,
with random, local dispersal that we take to be Gaussian
(although this should not affect the conclusions if the true dispersal distribution is not too fat-tailed).

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \cite{Kruuk1999,Barton2000a}.

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citep{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on numerous approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. 
%\alisa{I don't think I understand the independence from $\sigma$?}
%\plr{It gets set up at speed $\sigma$ over a distance proportional to $\sigma$.}

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $\sigma^2 \grad \log p(x,t)$ (i.e., it moves at speed $\sigma^2$ in the potential $-\log p$).
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x+r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$.

@@

Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ (same as previously described movement of a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$)or $r p$ (for $B$ to $A$).

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to compute expected clines in ancestry,
i.e., the expected proportion of individuals who inherit from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In other words,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

The resulting equation si similar to \eqref{eqn:q_pde} with a term added for branching,
which is written (omitting the $(x,t)$ for conciseness):
\begin{align}
    \begin{aligned} \label{eqn:g_pde}
        \dot g_A(a,b) 
        &= G^A g_A(a,b) + r(a,b) (1-p) (g_A(a,b)-g_B(a,b))
            \\ {} & \qquad 
            + (1-p) \left( \int_a^0 
                g_B(a,\theta) g_A(\theta,b)
                + \int_0^b g_A(a,\theta) g_B(\theta,b)
            d\theta \right)
            \\ {} & \qquad 
            + p \int_a^b {
                g_A(a,\theta) g_A(\theta,b) 
            } d\theta
            - (b-a) g_A(a,b)  
    \end{aligned} ,
\end{align}
where $\theta$ is the genomic position of a recombination event, 
the integral $\int_a^0$ is zero if $a>0$ and likewise $\int_0^b$ is zero if $b<0$;
and $r(a,b)$ is the distance from the segment to the selected site:
$r(a,b)=0$ if $ab<0$ and $r(a,b)=\min(|a|,|b|)$ otherwise.
The equation for $g_B(x,t;a,b)$ is identical after exchanging $A \leftrightarrow B$ and
$p \leftrightarrow (1-p)$.
The boundary conditions are $g_A(x,0;a,b)=1$ and $g_B(x,0;a,b)=0$.

@@

Each deme has exactly $N_d$ diploid individuals at the start of each generation.
At the start of each generation, every individual migrates to a new deme  offset by $\lfloor Z + 1/2 \rfloor$ from their original position,
where $Z$ is a Gaussian random variable with mean zero and variance $\sigma^2$.
(The mean displacement is zero, and when comparing to theory we compute $\sigma$ as the standard deviation of this distribution.)
Migrants past either end of the range remain at the terminal demes.
Then, 
fitnesses are computed (heterozygotes at the selected locus have fitness $1-s$; homozygotes have fitness 1),
and in each deme $N_d$ pairs of parents are chosen, with replacement,
with probability proportional to their fitness and selfing allowed.
Since migration is not conservative, demes may have no available parents;
in this case, parents are chosen from other demes with probability proportional to fitness
multiplied by $\exp(-x^2/\sigma^2)$, where $x$ is the distance to the other deme.
The next generation is formed by carrying out meiosis in each parent
and combining the resulting gametes such that each pair of parents leaves one descendant.
Meiosis results in alternating blocks of the gamete's chromsome
being inherited from the two parental chromosomes,
with the blocks separated by a Poisson(1) number of uniformly chosen recombination points along the chromosome,
and the order of the parental chromosomes chosen randomly.
The simulation software works by recording for each chromosome
a list of ancestry breakpoints and the index of the ancestor in the first generation
that the chromosome inherited that region of genome from;
genotyping at individual loci was performed by looking up the ancestor's genotype.

@@

\paragraph{The timescale over which linked clines relax:}
Although loci not under direct selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{Barton1986}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
We are interested, for samples across a cline $T$ units of time after secondary contact, how far away from the selected locus one has to look to see clines that resemble the genome-wide mean.

@@

The distribution of contiguous ancestry block lengths necessarily contains more information than allele frequency alone. We are interested specifically in the how tracts of ancestry surrounding the selected locus compares to the rest of the genome. 
The raw data -- ancestry assignments for a few individuals sampled from across space --
is shown in Figure~\ref{Fig:resistanceToIntrogression1000g},
showing distinct patterns across geography and the genome.

@@

It is possible that there is additional information contained in patterns of within-ancestry haplotype diversity throughout the hybrid zone. Looking at the number of ancestral haplotypes present in the population would provide informationa bout whether the migrant contribution is through relatively few, successful migrants, or through many migrants that each contribute relatively little.

@@

Of particular concern that stems from the above assumption is that our model ignores potential, and likely, sources of stochasticity arising from geographic structure and coalescence/pedigree structure. The latter can result in an underestimate of covariance in ancestry, as we have ignored additional sharing of ancestry through shared genealogy \cite{Liang2014}. With respect to genealogy, the broad agreement between our analytic predictions and simulations (which include an explicit pedigree structure), suggests that this assumption does not have a severe qualitative impact on our major results when populations are moderately high, especially for all but the most recently formed hybrid zones.

>>>>

While for much of the genome, gene flow will eventually erase 
both geographic clines in ancestry, and strong correlations in ancestry across linked markers, 
subsets of the genome can maintain stable clines in ancestry if selection against migrant, or hybrid, genotypes impedes gene flow across the hybrid zone \citep{Barton1979a}. 
% Variation in \yb{ the action of } selection against migrant alleles within the genome results in genome-wide heterogeneity in cline width, as the frequency of foreign ancestry surrounding targets of selection is reduced \citep{Barton1979}. 
% While these patterns of co-ancestry have been described in multi-locus systems, the \yb{expected length of ancestral tracts around selected loci in admixing populations remains poorly understood}\alisa{, especially in the geographic context of hybrid zones.}
\yb{While empirical patterns of the block length distribution in hybrid zones have rarely been used to inform the history of selection,}  \alisa{the availability of dense genotype data in combination with linkage information pave a way to utilize such patterns to better understand the processes operating in hybrid zones. } %GREAT SENTENCE!!
Here, we investigate expected patterns of co-ancestry \yb{ --- that is, the genomic scale at which ancestry strongly covaries between sites --- } across a hybrid zone in the presence of selection.

@@

We consider two isolated populations
--- labeled species $A$ and species $B$ ---   % yb: for consistency with the rest of the MS
that came into contact $T$ generations in the past.  
We will say that species $A$ was initially on the ``left'' of the zone of contact, which corresponds to spatial positions $x<0$.  
After contact, the populations live across continuous geographical space,
with random, local dispersal that we take to be Gaussian
(although this should not affect the conclusions if the true dispersal distribution is not too fat-tailed).

We model selection through a single, underdominant locus:
at this locus, there are two alleles, one corresponding to each of the ancestral populations.
Individuals who are heterozygous at this locus produce on average $1-s$ times as many offspring than either homozygote.
Although most selection in hybrid zones is likely more complex than this simple situation, 
previous work has shown that such models generate clines that are very similar to models of ecological selection or epistatic systems \citep{Kruuk1999,Barton2000a}.

@@

\paragraph{Establishment of the cline at the selected locus}
Deterministic theory predicts that after secondary contact,
the two alleles at the selected locus will form a stable cline \citep{Barton1979},
affecting neighboring loci as well.
First, we turn our attention to how the cline at the selected locus itself is formed.
Let $p(x,t)$ denote the frequency of the $A$ allele at spatial location $x$ and time $t$.  
Suppose that secondary contact occurred at time $t=0$, 
and that $p(x,0) = 1$ if $x<0$ and $p(x,0)=0$ if $x>0$.
% (In discrete computations, a deme at $x=0$ has $p(0,0)=1/2$.)
Let $\sigma^2$ be the mean squared distance between parent and offspring 
(i.e., the variance of the distribution of dispersal distance).
Assuming that: (i) alleles locally assort into diploids randomly, 
(ii) habitat and dispersal are homogeneous, and 
(iii) $s$ and $\sigma$ are small, then the commonly-used equation that $p$ solves is
\begin{align} \label{eqn:cline_pde}
    \dot p = \frac{\sigma^2}{2} \Delta p + s p (1-p) (2p-1) ,
\end{align}
where $\dot p(t,x)$ is the time derivative and $\Delta p$ is the Laplacian,
as derived in \citet{Bazykin1969}; for discussion of diploidy see \citet{christiansen1995genotypic}.
Although there is no known exact analytic solution to this equation, even in steady state, 
a good approximation to the equilibrium solution is that
$\lim_{t \to \infty} p(x,t) \approx (1+\tanh(-2x\sqrt{s}/\sigma))/2$ \citep{Bazykin1969}. 
This relies on numerous approximations, but the general conclusions should be quite robust: 
the stable cline has width of order $\sigma/\sqrt{s}$, and decays exponentially. 
Note that rescaling space and time by $\sigma/s$ and $1/s$ respectively in equation \eqref{eqn:cline_pde} results in a dimensionless equation,
implying that the cline establishes over a timescale of the order $1/s$. 
In other words, the cline is established through diffusion, which is slowed down somewhat by selection against heterozygotes. 
%\alisa{I don't think I understand the independence from $\sigma$?}
%\plr{It gets set up at speed $\sigma$ over a distance proportional to $\sigma$.}

@@

Making the same assumptions as for equation \eqref{eqn:cline_pde}, the derivation in Appendix \ref{apx:lineage_derivation} shows 
how the lineage of an $A$ allele moves as Brownian motion with speed $\sigma$
with drift $\sigma^2 \grad \log p(x,t)$ (i.e., it moves at speed $\sigma^2$ in the potential $-\log p$, note that we use ``drift'' as directional movement as in the terminology of diffusion, rather than its population genetic meaning).  % YB for clarity
% \alisa{note to self to read through appendix when written}
To see why this is true, note that the probability that the parent of an $A$ allele found at $x$
lived at position $x+r$ is proportional to $\P\{R=r\} p(x+r)$, where $R$ is the random dispersal distance;
and so the mean displacement from offspring to parent is $\E[R p(x+R)]/\E[p(x+R)]$.
% $$ = \E[ R ( p(x) + R p'(x) ) ]/ p(x) + O(R^3) $$
Using that $\E[R]=0$ and $\E[R^2]=\sigma^2$
and expanding $p()$ to first order about $x$ shows that this mean displacement is approximately $\sigma^2 p'(x)/p(x)$.

@@

Suppose we sample an allele today at a locus at recombination distance $r$ to the selected site.
We will follow its lineage back through time, using $\tau$ to denote ``generations ago'' to avoid confusion.
If $X_\tau$ is the geographic location of its ancestor $\tau$ generations ago,
and $Z_\tau$ is the identity of the selected allele that ancestor carried,
then we say that $X$ moves as a diffusion pushed by either $\log(p)$ or $\log(1-p)$ (same as previously described movement of a selected allele),
and that $Z$ jumps between $A$ and $B$ at rate either $r (1-p)$  (from $A$ to $B$) or $r p$ (for $B$ to $A$).

@@

\paragraph{Linked clines}
We can use this diffusion model for lineages to compute expected clines in ancestry,
i.e., the expected proportion of individuals who inherit ancestry from species $A$,
as a function of space, time, and position on the genome.
Precisely, we need the probability that 
an allele sampled $t$ generations after secondary contact at location $x$,
at recombination distance $r$ to a selected allele of type $z$,
is inherited from an individual of ancestry $A$,
where $z$ can be either $A$ or $B$.
We denote this probability $q_z(x,t,r)$.
In other words,
\begin{align}
    q_z(x,t,r) = \P^x \{Z_t = A\} .
\end{align}
(Recall that $Z$ depends implicitly on $r$.)
The description above implies that $q_z$ solves the following Kolmogorov backward equation:
\begin{align}
    \begin{aligned}  \label{eqn:q_pde}
    \dot q_A(x,t,r) 
            &= \sigma^2 \grad \log(p(x,t)) \cdot \grad q_A(x,t,r) 
                + \frac{\sigma^2}{2} \Delta q_A(x,t,r) 
            \\ &\qquad {} + 
                r (1-p(x,t))(q_B(x,t,r)-q_A(x,t,r))  \\
    \dot q_B(x,t,r) &= \sigma^2 \grad \log(1-p(x,t)) \cdot \grad q_B(x,t,r) 
            + \frac{\sigma^2}{2} \Delta q_B(x,t,r)
            \\ &\qquad {} + 
            r p(x,t) (q_A(x,t,r)-q_B(x,t,r))  ,
    \end{aligned} 
\end{align}
with boundary conditions $q_A(x,0,r)=1$ and $q_B(x,0,r)=0$, 
and where $\grad$ is the gradient, and $\Delta$ is the Laplacian.
%\alisa{this is fwd-in-time notation. should we change it to bwd-in-time?}
%\plr{This equation should be forwards in time, since it's the Kolmogorov backward equation
%applied to a backwards-time process!}

@@

The resulting equation is similar to \eqref{eqn:q_pde} with a term added for branching,
which is written (omitting the $(x,t)$ for conciseness):
\begin{align}
    \begin{aligned} \label{eqn:g_pde}
        \dot g_A(a,b) 
        &= G^A g_A(a,b) + r(a,b) (1-p) (g_A(a,b)-g_B(a,b))
            \\ {} & \qquad 
            + (1-p) \left( \int_a^0 
                g_B(a,\theta) g_A(\theta,b)
                + \int_0^b g_A(a,\theta) g_B(\theta,b)
            d\theta \right)
            \\ {} & \qquad 
            + p \int_a^b {
                g_A(a,\theta) g_A(\theta,b) 
            } d\theta
            - (b-a) g_A(a,b)  
    \end{aligned} ,
\end{align}
where $\theta$ is the genomic position of a recombination event, 
the integral $\int_a^0$ is zero if $a>0$ and likewise $\int_0^b$ is zero if $b<0$;
and $r(a,b)$ is the distance from the segment to the selected site:
$r(a,b)=0$ if $ab<0$ and $r(a,b)=\min(|a|,|b|)$ otherwise.
The equation for $g_B(x,t;a,b)$ is identical after exchanging $A \leftrightarrow B$ and
$p \leftrightarrow (1-p)$.
The boundary conditions are $g_A(x,0;a,b)=1$ and $g_B(x,0;a,b)=0$.

@@

Each deme has exactly $N_d$ diploid hermaphroditic individuals at the start of each generation.
At the start of each generation, every individual migrates to a new deme  offset by $\lfloor Z + 1/2 \rfloor$ from their original position,
where $Z$ is a Gaussian random variable with mean zero and variance $\sigma^2$.
(The mean displacement is zero, and when comparing to theory we compute $\sigma$ as the standard deviation of this distribution.)
Migrants past either end of the range remain at the terminal demes.
Then, 
fitnesses are computed (heterozygotes at the selected locus have fitness $1-s$; homozygotes have fitness 1),
and in each deme $N_d$ pairs of parents are chosen, with replacement,
with probability proportional to their fitness and selfing allowed.
Since migration is not conservative, demes may have no available parents;
in this case, parents are chosen from other demes with probability proportional to fitness
multiplied by $\exp(-x^2/\sigma^2)$, where $x$ is the distance to the other deme.
The next generation is formed by carrying out meiosis in each parent
and combining the resulting gametes such that each pair of parents leaves one descendant.
Meiosis results in alternating blocks of the gamete's chromsome
being inherited from the two parental chromosomes,
with the blocks separated by a Poisson(1) number of uniformly chosen recombination points along the chromosome,
and the order of the parental chromosomes chosen randomly.
The simulation software works by recording for each chromosome
a list of ancestry breakpoints and the index of the ancestor in the first generation
that the chromosome inherited that region of genome from;
genotyping at individual loci was performed by looking up the ancestor's genotype.

@@

\paragraph{The timescale over which linked clines relax:}
Although loci not under direct selection can in principle spread across the cline unimpeded,
in practice it can take quite some time for even unlinked neutral loci to homogenize,
due to the decreased fitness of heterozygotes \citep{Barton1986}
and the relative slowness of diffusive movement \citep{Sedghifar2015}.
We wish to know, for samples across a cline $T$ units of time after secondary contact, how far away from the selected locus one has to look to see clines that resemble the genome-wide mean.

@@

The distribution of contiguous ancestry block lengths necessarily contains more information than allele frequency alone. We are interested specifically in how the tracts of ancestry surrounding the selected locus compares to the rest of the genome. 
The raw data -- ancestry assignments for a few individuals sampled from across space --
is shown in Figure~\ref{Fig:resistanceToIntrogression1000g},
showing distinct patterns across geography and the genome.

@@

It is possible that there is additional information contained in patterns of within-ancestry haplotype diversity throughout the hybrid zone. Looking at the number of ancestral haplotypes present in the population would provide information about whether the migrant contribution is through relatively few, successful migrants, or through many migrants that each contribute relatively little.

@@

Of particular concern that stems from the above assumption is that our model ignores potential, and likely, sources of stochasticity arising from geographic structure and coalescence/pedigree structure. The latter can result in an underestimate of covariance in ancestry, as we have ignored additional sharing of ancestry through shared genealogy \citep{Liang2014}. With respect to genealogy, the broad agreement between our analytic predictions and simulations (which include an explicit pedigree structure), suggests that this assumption does not have a severe qualitative impact on our major results when populations are moderately high, especially for all but the most recently formed hybrid zones.
